\chapter{Definitions}
\label{c:definitions}

Pattern discovery is a research area aiming to discover unknown patterns in a given set of data structures that are frequent and interesting according to some measure. In this chapter we formally define necessary terms used in this thesis.

\section{Sequence and Dataset}

We use $\Sigma$ to denote the set of tokens in the dataset, an \emph{alphabet}. The \emph{size} of the alphabet is $|\Sigma|$. \emph{Tokens} can be numbers, letters, words or sentences or any other fixed element.

Any sequence $S=a_1 a_2 ... a_n, \forall a_i \in \Sigma$ is called a \emph{sequence} over the alphabet $\Sigma$. If the length of the string is $0$, it is called an empty sequence or $\epsilon$.

\begin{exmp}
\R{ACGTGCCATC} is a sequence over $\Sigma = \{\R{A}, \R{C}, \R{G}, \R{T}\}$.
\end{exmp}

A \emph{dataset} is a collection of sequences.

\begin{exmp}
In a document, sentences can be considered as a \emph{dataset}, where a single sentence is a \emph{sequences} and each word is a \emph{token} in the alphabet. The text \R{This is some example. This is an other example.} has sequences \{ [\R{This is an example}], [\R{This is an other example}] \} and the alphabet is $\Sigma =$ \{\R{this}, \R{is}, \R{an}, \R{example}, \R{other}\}.	
\end{exmp}

\section{Pattern}

Our aim is to discover repetitive and common structures in data. We call such structures \emph{patterns}. One generic way to define a \emph{pattern} is as a set of all the sub-structures it represents. This means that we can say whether some data sub-structure is represented by a pattern.

The \emph{pattern structure} is usually dependent on the data-structures which it represents. For example, sequence patterns are usually represented as sequences, graph patterns are represented as graphs; but in some situations sequence patterns could also be represented as graphs.

We denote the set of structures that a pattern structure $p$ defines as $all(p)$. If $\alpha \in all(p)$, where $\alpha$ is a structure then we say that the structure $\alpha$ \emph{matches exactly} the pattern $p$. We say that $\alpha$ \emph{matches} $p$ if any structure from $all(p)$ is a sub-structure of $\alpha$.

In this thesis we only consider sequential pattern structures and use \emph{pattern} in the meaning \emph{sequential pattern structure}. One very common way of describing patterns is by using regular expressions\cite{KleeneRegularSets,RegularExpressions}. In this thesis we use regular expressions to describe the patterns and introduce the regular expression syntax where needed.

\emph{Pattern size} is the length of the pattern sequence.

\begin{exmp}
\R{.[AT]} is a pattern of size 2 and denotes a set 
\{ \R{AA}, \R{AT}, \R{CA}, \R{CT}, \R{GA}, \R{GT}, \R{TA}, \R{TT}\}; it matches \R{CCTC} and exactly matches \R{AT}.	
\end{exmp}

We denote the set of all pattern $p$ \emph{matches} in a dataset $D$ as $matches(p, D) = \{ match(p, \alpha) | \alpha \in D, matches(p, \alpha) \}$.

\section{Query}

We need to somehow understand where a given pattern $p$ is located in the dataset $D$. This compound structure $q = <D, p, matches(p, D)>$ is called a \emph{query}.

\begin{exmp}
Let our dataset be $D = [ \text{\R{ACGT}}, \text{\R{TXCGA}} ]$ and our pattern be $p = \text{\R{C.}}$. The corresponding query is $<D, p, \{ [1,3], [2,4]\}>$, which means that in the sequence 1, pattern $p$ ends at position 3, and in the sequence 2, the pattern ends at position 4.
\end{exmp}

The position representation is largely dependent on the dataset and pattern representation, but irrelevant for this generic algorithm.

\subsection{Query features}

When we talk about how "interesting" a pattern is, we are actually evaluating the query, since the pattern requires a context where it can be "interesting".

Queries can have different properties: length, number of matches in the dataset, pattern textual representation etc. Such properties can be represented by a function that takes a query as an input and returns the property. Formally a \emph{query feature} is a function $f: Query \mapsto Any$.

We also need to see how "interesting" one query is compared to the others. \emph{Query interestingness} is a function $f: Query \mapsto Value$ where the $Value$-s are well-ordered. This gives a measure to compare two different queries. We can often represent such interestingness measures as a real number.

We should also be able to somehow specify criteria for query. \emph{Query filter} is a function $f: Query \mapsto Boolean$ and shows whether the query matches the criteria.

\begin{exmp}
Pattern occurrences in a document can be considered an interestingness measure. Whether query pattern is at least 3 tokens long is a query filter.
\end{exmp}

\section{Pool}

\emph{Pool} is an abstract data type for a storing the queries. The only operations that pool must provide is "push", for adding a query, and "pop", for getting a query.

\begin{exmp}
Stacks and queues both satisfy the pool requirement. We could also define pools that store the queries on the disk; it could internally pack or reorder the queries for performance reasons.
\end{exmp}

\section{Pattern Discovery}

In this thesis, \emph{pattern discovery} is defined as a process for finding, according to a query interestingness measure, the most interesting subset of sequential patterns that meet certain criteria in a sequence dataset.

\begin{exmp}
Let our research problem be "Finding the most common nucleotide patterns, which are at least 3 nucleotides long, from a shotgun sequencing output.".\emph{Most common} defines our interestingness measure. \emph{At least 3 nucleotides} is the criteria for selecting a subset of patterns. \emph{Sequencing output} is used as our dataset and \emph{nucleotides} define how the sequence looks like.
\end{exmp}